From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-17 paper.tex
Rewritten introduction, re-organized the paper one more time
Starts to look better IMHO :-)

In order to build intelligent agents, it is necessary to equip them with a range
of perceptual capabilities: while face and object tracking and
recognition~\cite{Zhao:2003, Jafri:2014}, path planning~\cite{Galceran:2013},
speech recognition~\cite{brick2007incremental} or task
learning~\cite{calinon2007learning} are amongst the current major research
directions, existing literature tends to evaluate each of these algorithms in
their own metric space, without considering the overall interaction quality.
Along this line, Anzalone~\etal~\cite{anzalone} have recently argued in favor of
such a global evaluation, and propose to assess these algorithms in term of
their capability to obtain the desired effect in a human-robot interaction
context, and they correspondingly propose metrics build around the measurement
of \emph{engagement} as indicator of the quality of the experience.

Engagement describes a cognitive, affective (specifically intrinsic motivation),
and behavioral state of interaction with a computer application that ``makes the
user want to be there" \cite{OBrien:2010}. Engaging interactions were thought to
involve intrinsic interest, interactivity, perceived control and choice,
functionality, motivation and attention.

\paragraph{Motivation: Impact on Learning}

Education and learning environments are interesting application fields for
research on attention assessment, since sustaining children's engagement over
time is a key variable to ensure effective learning (see for
instance~\cite{Umbach}), and they indeed motivate this research. In fact, the
use of robots in education comes with the potential to engage the child in
meta-cognition through the learning by teaching paradigm, wherein a student
takes the role of a teacher and experiences stronger educational benefits as a
result (such as in \cite{Palinscar1984}).

While~\cite{zhao2012learning} recommends the development of a teachable agent
that is actively, rather than passively, engaged in the learning by teaching
interaction, the increased learning gain results attained
in~\cite{okita2006observation} were done so even with the pupil being taught
having been instructed to not contribute much information. This suggests that it
may not be necessary, from a learning by teaching point of view, for the
teachable agent to participate actively in the interaction to incite educational
benefits. However, from an engagement point of view, this may not necessarily be
true.

The goal of this research is therefore to \textbf{evaluate the quality of an
interaction} using the \textbf{visual focus of attention as an engagement
indicator}. We validate our approach with field experiments conducted in the
educational context of the CoWriter project~\cite{Hood:2015}, where our
technique, being based on a regular, low-end RGB camera, avoids the use of
otherwise typically invasive technologies such as eyetrackers, that would
negatively impact the natural of the interaction.

\subsection*{Attention and Engagement: Related Work}

\paragraph{Head Tracking and Visual Attention}

@@

\paragraph{Engagement}

The notion of engagement has actively been studied in a diverse set of domains,
and several variables and social signals have been proposed in the literature to
quantify and study it in social interactions.

The evaluation of response times to provide an estimation of the user's
engagement based on the correctness has been proposed in a learning context
\cite{Beck} before. However, several aspects like the inter-subject variability
or the environment complexity make it hardly applicable in a child-robot
interaction. Additionally, other techniques \cite{Chipman07postureas} like the
use of the weight distribution using a pressure sensitive chair has reported
results. But it is in the computer vision field where major contributions have
been done.

In~\cite{peters2010investigating} approach, a model of user's interest and
engagement with a virtual agent displaying shared attention behavior, by using
eye gaze and head direction information is proposed. Similarly, in
\cite{nakano2010estimating} an approach to estimate the user's conversational
engagement with a conversational agent based on analysis of gaze patterns was
implemented. Other significant work \cite{Castellano:2009} focus on feature
evaluation related to the user's non-verbal behavior, the task and the
companion's affective reactions are identified to predict the children's level
of engagement using a Bayesian model.

Some few studies have proposed approaches for the detection of engagement with a
robot. In \cite{Rich:2010} a computational model based on the recognition of
connection events such as directed gaze, mutual facial gaze is proposed. In this
context, some other \cite{Sanghvi:2011} focus only on the back or trunk posture
as a determined factor for the assessment. A recent study \cite{anzalone} with
social robots in face to face scenarios proposed a set of metrics based on
non-verbal cues but manifesting some possible limitations in long-term
scenarios.

In~\cite{baxter2014tracking}, Baxter~\etal posit that the measure of the
direction and timing of gaze in child-robot interactions is a proxy for
engagement and attribution of social agency, However, they conduct these
measures as post-hoc analyses. This work suggests a path forward for real-time
assessment of engagement.

In this research, we address the problem of tracking the children's engagement
based on their VFoA and current performed task during a long term child-robot
interaction in a learning context, first. And second, we provide qualitative and
quantitative evaluation of the outcome provided using a first person RGB sensor
approach. Moreover, this monitoring activity has been supporting our research in
child-robot interaction for learning and instruction purposes.

\subsection*{Overview of the Article}

\paragraph{Experimental Validation of the Approach}

\fixme{TBD}

\paragraph{From Attention to Engagement}

\fixme{TBD}

@@

\begin{table}[h!]
	\centering
	\caption{Tracking accuracy in face-to-face scenario}
	\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 \\ \hline
	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 \\ \hline
	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 \\ \hline
	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 \\ \hline
	\end{tabular}
	\label{tab:results}
\end{table}

@@

\section{Discussion and Conclusion}
\subsection{From Attention to Engagement Assessment}

@@

\subsection{Conclusion}

>>>>

Building capable social agents requires to equip them with a range
of perceptual capabilities: while face and object tracking and
recognition, path planning, speech recognition or task learning are some of the current major research
directions, existing literature tends to evaluate each of these algorithms in
their own metric space, without considering the interaction quality at a global
level: Anzalone~\etal~\cite{anzalone} have recently argued in favor of
such a global evaluation, and they propose to assess these algorithms in term of
their capability to \emph{obtain the desired effect} in a human-robot interaction
context. They correspondingly propose metrics build around the measurement
of \emph{engagement} as indicator of the quality of the experience.

``Engagement'', the cognitive, affective and behavioral state of interaction
with a computer application that ``makes the user want to be
there"~\cite{OBrien:2010}, has actively been studied in a diverse set of
domains, and specifically in robotics, several variables and social signals have been proposed in the
literature to quantify it. A recent review of these is in~\cite{ivaldi2015towards}.

%In a learning context, the evaluation of response times in relation with
%correctness has been proposed~\cite{Beck} to provide estimate the user's 
%engagement, with however large inter-subject and environment variability.
%Biometric measurements have also been proposed, like the use the weight
%distribution of the user on a pressure-sensitive
%chair~\cite{Chipman07postureas}.

For instance, \cite{Castellano:2009} proposes to predict children's level of
engagement by integrating in a Bayesian model non-verbal cues (gaze and smiles)
with the current state of the interaction. While they report high level of
accuracy, their approach requires post-hoc video annotations, and is not
applicable to on-line engagement assessment. Similarly,
Baxter~\etal~\cite{baxter2014tracking} posit that the measure of the direction
and timing of gaze in child-robot interactions is a proxy for engagement and
attribution of social agency. However, they also conduct these measures as
post-hoc analyses.  \cite{peters2010investigating} model the user's interest and
engagement with a virtual agent by tracking eye gaze and head direction.
Similarly, \cite{ishii2011combining} estimates the user's engagement with a
conversational agent based on the analysis of gaze patterns.
In~\cite{Rich:2010}, a computational model based on the recognition of
connection events such as directed gaze, mutual facial gaze is proposed. Not
relying on gaze, \cite{Sanghvi:2011} focus only on the back or trunk posture as
a determining factor for the assessment. A recent study~\cite{anzalone} with
social robots in face-to-face scenarios proposed a set of metrics based on
non-verbal cues but manifesting some possible limitations in long-term
scenarios.

The variety of these approaches reflects the fact that \emph{engagement} remains
a broad concept, fairly ill-defined and thus difficult to operationalize.
Therefore, instead of introducing ``yet another metric of engagement'', we
propose in this paper to introduce the more specific concept of
``\emph{with-me-ness}'': to what extend the human remains ``with me'', the
robot, during our interaction.

We assess this level of ``with-me-ness'' by combining real-time estimation of the
attentional focus of the human with the expected, {\it a priori} focus
requested by the task, as decided by the robot.

As such, this article is organized in three sections: we first present a novel
method for on-line estimation of the focus of attention based on fast 6D head
pose estimation. We then present an extensive validation of this approach based on
a real-world field experiment with 6 children. We finally show how to compute in
real-time ``with-me-ness'' to provide a new \emph{in-the-moment} measurement of
the quality of interaction, and compare levels computed by the robot with a
ground truth based on manual post-hoc annotations.

%\paragraph{Motivation: Impact on Learning}
%
%Education and learning environments are interesting application fields for
%research on attention assessment, since sustaining children's engagement over
%time is a key variable to ensure effective learning (see for
%instance~\cite{Umbach}), and they indeed motivate this research. In fact, the
%use of robots in education comes with the potential to engage the child in
%meta-cognition through the learning by teaching paradigm, wherein a student
%takes the role of a teacher and experiences stronger educational benefits as a
%result (such as in \cite{Palinscar1984}).
%
%While~\cite{zhao2012learning} recommends the development of a teachable agent
%that is actively, rather than passively, engaged in the learning by teaching
%interaction, the increased learning gain results attained
%in~\cite{okita2006observation} were done so even with the pupil being taught
%having been instructed to not contribute much information. This suggests that it
%may not be necessary, from a learning by teaching point of view, for the
%teachable agent to participate actively in the interaction to incite educational
%benefits. However, from an engagement point of view, this may not necessarily be
%true.
%
%The goal of this research is therefore to \textbf{evaluate the quality of an
%interaction} using the \textbf{visual focus of attention as an engagement
%indicator}. We validate our approach with field experiments conducted in the
%educational context of the CoWriter project~\cite{Hood:2015}, where our
%technique, being based on a regular, low-end RGB camera, avoids the use of
%otherwise typically invasive technologies such as eyetrackers, that would
%negatively impact the natural of the interaction.

\subsection{Related Work}

@@

@@

\begin{table}[h!]
	\centering
    \caption{Tracking accuracy in face-to-face scenario\fixme{std dev needed!}}
	\begin{tabular}{|c|c|c|c|c|c|c|c|} \hline
	\small Subject & 1 & 2 & 3 & 4 & 5 & 6 & Avg\\ \hline
	\small Track(\%) & \small88.2 & \small83.5 & \small90.5 & \small83.1 & \small87.9 & \small85.0 & \small86.4 \\ \hline
	\begin{tabular}{@{}c@{}}\small Real vs \\ \small expected(\%)\end{tabular} & \small92.6 & \small88.8  & \small95.8 & \small93.1 & \small96.5 & \small89.6 & \small92.7 \\ \hline
	\begin{tabular}{@{}c@{}}\small Real vs \\ \small captured(\%)\end{tabular} & \small61.1 & \small85.5 & \small80.8 & \small62.2 & \small63.4 & \small83.9 & \small72.8 \\ \hline
	\begin{tabular}{@{}c@{}}\small Cosine \\ \small affinity \end{tabular} & \small0.73 & \small0.82 & \small0.87 & \small0.73 & \small0.77 & \small0.89 & \small0.8 \\ \hline
	\end{tabular}
	\label{tab:results}
\end{table}

@@

\section{With-me-ness}

@@

\section{Discussion and Conclusion}
